---
title: "Malaria detector"
output: html_document
date: "`r Sys.Date()`"
---

```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)
```
```{r}
# Load the data from Url
url <- "https://raw.githubusercontent.com/HackBio-Internship/public_datasets/main/R/lancet_malaria.txt"
Malaria <- read.table(url, header = TRUE, sep = "\t") 
```

```{r}
data <- Malaria

# Removing rows with NA values
malaria_cleaned <- na.omit(data)

# check the structure of the data
str(malaria_cleaned)

```

```{r}
# Scatter plot of PCR % against Microscopy %
ggplot(malaria_cleaned, aes(x = X..Microscopy..Positive, y = X..PCR..Positive)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +  # Add a 1:1 line
  labs(x = "Microscopy % Positive", y = "PCR % Positive", title = "PCR % vs Microscopy %")
```
This is a combined scatter plot of PCR % Positive against Microscopy % Positive across all regions.
The 1:1 dotted line serves as a reference. Points below this line indicate higher PCR detection rates than microscopy.
There is a clustering of points below the 1:1 line, especially at lower microscopy percentages. This pattern suggests that across the dataset, PCR consistently detects more cases at lower infection levels, indicating widespread submicroscopic infections in the sample.

```{r}
# Add prevalence ratio to the dataframe
data <- malaria_cleaned %>%
  mutate(Prevalence_Ratio = Microscopy.N.Positive / PCR.N.Positive)

# Check the updated dataframe
head(data)

```
```{r}
# Scatter plot of PCR % vs Microscopy % for four global regions
ggplot(data, aes(x = X..Microscopy..Positive, y = X..PCR..Positive)) +
  geom_point() +
  facet_wrap(~ Global.Region) +  # Facet by global region
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +  # Add a 1:1 line
  labs(x = "Microscopy % Positive", y = "PCR % Positive", title = "PCR % vs Microscopy % by Region")

```
This scatter plot shows PCR % Positive against Microscopy % Positive, separated into two panels for East Africa and West Africa.
East Africa:
There are many points below the 1:1 dotted line, indicating that PCR detects more positive cases than microscopy at lower percentages, suggesting a significant number of submicroscopic infections.
This is especially visible at lower microscopy percentages, where the PCR percentages are relatively higher.
West Africa:
There is still a noticeable presence of points below the 1:1 line, but the distribution appears slightly closer to the line than in East Africa.
This suggests that while submicroscopic infections are present, the difference between PCR and microscopy detection rates may be less pronounced in West Africa compared to East Africa.
```{r}
# Boxplot of prevalence ratio by global region
ggplot(data, aes(x = Global.Region, y = Prevalence_Ratio)) +
  geom_boxplot() +
  labs(x = "Global Region", y = "Prevalence Ratio", 
       title = "Prevalence Ratio by Global Region")

```
This boxplot shows the distribution of the prevalence ratio (Microscopy Positive / PCR Positive) for East Africa and West Africa.
East Africa:
The boxplot has a wider range with some outliers, indicating variability in the prevalence ratio.
The median value appears to be lower than West Africa, which aligns with the observation that PCR tends to detect more positives than microscopy in this region.
West Africa:
The distribution is narrower and centered around a slightly higher median than East Africa.
This suggests that the prevalence ratio in West Africa is closer to 1, meaning that microscopy results are more aligned with PCR in this region.


conclusion:
Overall, PCR tends to identify more infections than microscopy, particularly at lower prevalence rates, reinforcing the presence of submicroscopic infections in the population.
East Africa shows a higher density of submicroscopic infections as indicated by the larger gap between PCR and microscopy rates at lower microscopy values.
The lower prevalence ratio in East Africa further supports the idea that there is a higher density of submicroscopic infections in East Africa compared to West Africa
